---
title: "DGE"
format: html
editor: visual
---

```{r}
#if (!require("BiocManager")) install.packages("BiocManager")
#BiocManager::install(c("tximport", "DESeq2"))
#install.packages(c("ggplot2", "pheatmap", "RColorBrewer", "ggrepel", "dplyr"))
```

```{r}
# ============================================================
# De novo RNAseq DGE Analysis
# Flutamide experiment — Frog
# tximport + DESeq2
# ============================================================

library(tximport)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(ggrepel)
library(dplyr)
```

# 1. File paths

```{r}
# ============================================================
# 1. File paths
# ============================================================

# Set base directory
base_dir <- "~/Desktop/Thesis"

# Sample names
sample_names <- c(
  "C103-2B", "C103-3B", "C103-4B", "C103-4H", "C103-5B", "C103-5H",
  "C45-1B",  "C45-1H",  "C45-2B",  "C45-2H",  "C45-3B",  "C45-4B",
  "C45-4H",  "C45-5B",  "C45-5H",
  "F103-1B", "F103-1H", "F103-2B", "F103-2H", "F103-3B", "F103-3H",
  "F103-4B", "F103-5B", "F103-5H"
)

# Paths to quant.sf files
files <- file.path(base_dir, "salmon_quant_v1", sample_names, "quant.sf")
names(files) <- sample_names

# Verify all files exist
stopifnot(all(file.exists(files)))
cat("All 24 quant.sf files found\n")
```

```{r}
#print(head(tx2gene))
```


# 2. Load transcript-to-gene map
```{r}
# ============================================================
# 2. Load transcript-to-gene map
# ============================================================

#tx2gene <- read.table(
#  file.path(base_dir, "trinity_out_dir.Trinity.fasta.gene_trans_map"),
#  header = FALSE,
#  col.names = c("gene_id", "transcript_id")
#)

# tximport expects transcript first, gene second
#tx2gene <- tx2gene[, c("transcript_id", "gene_id")]

#cat("tx2gene dimensions:", dim(tx2gene), "\n")
#head(tx2gene)
```

```{r}
tx2gene <- read.table(
  "~/Desktop/Thesis/trinity_out_dir.Trinity.fasta.gene_trans_map",
  header = FALSE,
  col.names = c("gene_id", "transcript_id")
)

# tximport expects transcript first, gene second
tx2gene <- tx2gene[, c("transcript_id", "gene_id")]

head(tx2gene)
dim(tx2gene)
```


# 3. Import with tximport
```{r}
# ============================================================
# 3. Import with tximport
# ============================================================

txi <- tximport(
  files,
  type = "salmon",
  tx2gene = tx2gene,
  ignoreTxVersion = FALSE
)

cat("tximport complete\n")
cat("Dimensions of count matrix:", dim(txi$counts), "\n")
```

# 4. Build sample metadata (colData)
```{r}
# ============================================================
# 4. Build sample metadata (colData)
# ============================================================

colData <- data.frame(
  sample    = sample_names,
  treatment = factor(ifelse(grepl("^C", sample_names), "control", "flutamide"),
                     levels = c("control", "flutamide")),
  timepoint = factor(ifelse(grepl("45", sample_names), "day45", "day103"),
                     levels = c("day45", "day103")),
  tissue    = factor(ifelse(grepl("B$", sample_names), "body", "head"),
                     levels = c("body", "head")),
  row.names = sample_names
)

# Create group factor combining treatment and timepoint
colData$group <- factor(
  paste(colData$treatment, colData$timepoint, sep = "_"),
  levels = c("control_day45", "control_day103", "flutamide_day103")
)

cat("\nSample metadata:\n")
print(colData)
```

# 5. Create DESeq2 object
```{r}
# ============================================================
# 5. Create DESeq2 object
# ============================================================

dds <- DESeqDataSetFromTximport(
  txi,
  colData = colData,
  design  = ~ tissue + group
)

cat("\nDESeq2 object created\n")
cat("Number of genes:", nrow(dds), "\n")
```

# 6. Pre-filtering
```{r}
# ============================================================
# 6. Pre-filtering
# ============================================================

# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
cat("Genes after filtering:", nrow(dds), "\n")
```

# 7. Run DESeq2
```{r}
# ============================================================
# 7. Run DESeq2
# ============================================================

dds <- DESeq(dds)
cat("\nDESeq2 complete\n")
```

# 8. Sample-level QC — PCA and clustering
```{r}
# ============================================================
# 8. Sample-level QC — PCA and clustering
# ============================================================

# Variance stabilizing transformation for visualization
vst <- vst(dds, blind = TRUE)

# PCA plot
pca_data <- plotPCA(vst, intgroup = c("group", "tissue"), returnData = TRUE, ntop = nrow(vst))
percent_var <- round(100 * attr(pca_data, "percentVar"))

pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2,
                                  color = group,
                                  shape = tissue,
                                  label = name)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3) +
  xlab(paste0("PC1: ", percent_var[1], "% variance")) +
  ylab(paste0("PC2: ", percent_var[2], "% variance")) +
  scale_color_manual(values = c("control_day45"   = "#2166AC",
                                "control_day103"  = "#92C5DE",
                                "flutamide_day103" = "#D6604D")) +
  theme_bw() 

ggsave(file.path(base_dir, "figures/PCA_all_samples.png"),
       pca_plot, width = 8, height = 6)
cat("PCA plot saved\n")

# Sample distance heatmap
sample_dists <- dist(t(assay(vst)))
sample_dist_matrix <- as.matrix(sample_dists)

annotation_col <- data.frame(
  group   = colData$group,
  tissue  = colData$tissue,
  row.names = rownames(colData)
)

pdf(file.path(base_dir, "figures/sample_distance_heatmap.pdf"),
    width = 10, height = 8)
pheatmap(sample_dist_matrix,
         clustering_distance_rows = sample_dists,
         clustering_distance_cols = sample_dists,
         annotation_col = annotation_col,
         color = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
         main = "Sample distance heatmap")
dev.off()
cat("Sample distance heatmap saved\n")

png(file.path(base_dir, "figures/sample_distance_heatmap.png"),
    width = 1000, height = 800, res = 150)

pheatmap(sample_dist_matrix,
         clustering_distance_rows = sample_dists,
         clustering_distance_cols = sample_dists,
         annotation_col = annotation_col,
         color = colorRampPalette(rev(brewer.pal(9, "Blues")))(255))

dev.off()

cat("Sample distance heatmap saved\n")
```
# no of PC
```{r}
# Your pca_data only has PC1 & PC2 by default — use prcomp directly
pca_full <- prcomp(t(assay(vst)))
summary(pca_full)               # shows variance explained for ALL PCs
```


```{r}
# Max PCs = min(nsamples - 1, ngenes)
ncol(vst)   # number of samples
nrow(vst)   # number of genes
# Max meaningful PCs = n_samples - 1
```
```{r}
# Replicate exactly what plotPCA does internally
rv <- rowVars(assay(vst))
select <- order(rv, decreasing = TRUE)[1:500]
pca_full <- prcomp(t(assay(vst)[select, ]))
summary(pca_full)   # should now match 69% / 16%
```

# 9. Differential expression — planned contrasts
```{r}
# ============================================================
# 9. Differential expression — planned contrasts
# ============================================================

# Make figures directory
dir.create(file.path(base_dir, "figures"), showWarnings = FALSE)
dir.create(file.path(base_dir, "results"), showWarnings = FALSE)

# Function to run contrast, save results and volcano plot
run_contrast <- function(dds, contrast, name, base_dir) {

  res <- results(dds,
                 contrast = contrast,
                 alpha = 0.05)

  res <- res[order(res$padj), ]
  summary(res)

  # Save results table
  write.csv(as.data.frame(res),
            file.path(base_dir, "results", paste0(name, "_DESeq2_results.csv")))

  # Volcano plot
  res_df <- as.data.frame(res) %>%
    mutate(
      significant = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1,
      direction   = case_when(
        significant & log2FoldChange > 1  ~ "up",
        significant & log2FoldChange < -1 ~ "down",
        TRUE ~ "ns"
      )
    )

  volcano <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj),
                                 color = direction)) +
    geom_point(alpha = 0.5, size = 1) +
    scale_color_manual(values = c("up" = "#D6604D",
                                  "down" = "#2166AC",
                                  "ns" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
    theme_bw() +
    ggtitle(name) +
    xlab("log2 Fold Change") +
    ylab("-log10 adjusted p-value")

  ggsave(file.path(base_dir, "figures", paste0(name, "_volcano.pdf")),
         volcano, width = 7, height = 5)

  cat("\n---", name, "---\n")
  cat("Total DEGs (padj < 0.05, |LFC| > 1):",
      sum(!is.na(res$padj) & res$padj < 0.05 & abs(res$log2FoldChange) > 1), "\n")
  cat("Upregulated:", sum(!is.na(res$padj) & res$padj < 0.05 & res$log2FoldChange > 1), "\n")
  cat("Downregulated:", sum(!is.na(res$padj) & res$padj < 0.05 & res$log2FoldChange < -1), "\n")

  return(res)
}

# Contrast 1: Flutamide effect (F103 vs C103) — PRIMARY
res_flutamide <- run_contrast(
  dds,
  contrast = c("group", "flutamide_day103", "control_day103"),
  name     = "flutamide_effect_F103vsC103",
  base_dir = base_dir
)

# Contrast 2: Timepoint effect in controls (C103 vs C45)
res_timepoint <- run_contrast(
  dds,
  contrast = c("group", "control_day103", "control_day45"),
  name     = "timepoint_effect_C103vsC45",
  base_dir = base_dir
)

# Contrast 3: Flutamide vs baseline (F103 vs C45)
res_baseline <- run_contrast(
  dds,
  contrast = c("group", "flutamide_day103", "control_day45"),
  name     = "flutamide_vs_baseline_F103vsC45",
  base_dir = base_dir
)

# Contrast 4: Tissue effect (body vs head)
res_tissue <- run_contrast(
  dds,
  contrast = c("tissue", "body", "head"),
  name     = "tissue_effect_bodyvshead",
  base_dir = base_dir
)
```

# 10. Heatmap of top DEGs
```{r}
# ============================================================
# 10. Heatmap of top DEGs
# ============================================================

# Top 50 DEGs from flutamide contrast
top_genes <- rownames(res_flutamide)[1:50]

mat <- assay(vst)[top_genes, ]
mat <- mat - rowMeans(mat)

annotation_col <- data.frame(
  group  = colData$group,
  tissue = colData$tissue,
  row.names = rownames(colData)
)

pdf(file.path(base_dir, "figures/heatmap_top50_flutamide_DEGs.pdf"),
    width = 12, height = 10)
pheatmap(mat,
         annotation_col = annotation_col,
         show_rownames  = FALSE,
         show_colnames  = TRUE,
         color = colorRampPalette(c("#2166AC", "white", "#D6604D"))(255),
         main  = "Top 50 DEGs — flutamide effect")
dev.off()
cat("Heatmap saved\n")
```


# 11. Save DESeq2 object
```{r}
# ============================================================
# 11. Save DESeq2 object
# ============================================================

saveRDS(dds, file.path(base_dir, "results/dds_all_samples.rds"))
saveRDS(vst, file.path(base_dir, "results/vst_all_samples.rds"))
cat("\nDESeq2 object saved\n")
cat("\nAnalysis complete!\n")
cat("Results in:", file.path(base_dir, "results"), "\n")
cat("Figures in:", file.path(base_dir, "figures"), "\n")
```




# Diamond
```{r}
diamond <- read.delim(
  "diamond_swissprot.tsv",
  header = FALSE,
  sep = "\t",
  stringsAsFactors = FALSE
)

colnames(diamond) <- c(
  "qseqid","sseqid","pident","align_len","mismatch","gapopen",
  "qstart","qend","sstart","send","evalue","bitscore","stitle"
)

# remove peptide suffix: .p1, .p2, etc.
diamond$transcript_id <- sub("\\.p[0-9]+$", "", diamond$qseqid)
```


```{r}
gtm <- read.delim(
  "trinity_out_dir.Trinity.fasta.gene_trans_map",
  header = FALSE,
  sep = "\t",
  stringsAsFactors = FALSE
)

colnames(gtm) <- c("gene_id", "transcript_id")

diamond2 <- merge(diamond, gtm, by = "transcript_id", all.x = TRUE)

# keep best hit per gene by bitscore, then by evalue
diamond2 <- diamond2[order(diamond2$gene_id, -diamond2$bitscore, diamond2$evalue), ]
diamond_gene <- diamond2[!duplicated(diamond2$gene_id), ]

# optional: pull a cleaner gene symbol from GN=
diamond_gene$gene_symbol <- ifelse(
  grepl(" GN=", diamond_gene$stitle),
  sub(".* GN=([^ ]+).*", "\\1", diamond_gene$stitle),
  NA
)
```


```{r}
write.csv(
  diamond_gene,
  "diamond_gene_annotation_v1.csv",
  row.names = FALSE
)
```

# Results 
# Merge flutamide_effect_F103vsC103_DESeq2_results
```{r}
res <- read.csv(
  "results/flutamide_effect_F103vsC103_DESeq2_results.csv",
  row.names = 1,
  stringsAsFactors = FALSE
)

res$gene_id <- rownames(res)

annot <- read.csv(
  "diamond_gene_annotation_v1.csv",
  stringsAsFactors = FALSE
)

res_annot <- merge(res, annot, by = "gene_id", all.x = TRUE)

write.csv(
  res_annot,
  "flutamide_effect_F103vsC103_DESeq2_results_annotated.csv",
  row.names = FALSE
)
```

```{r}
sum(!is.na(diamond_gene$gene_symbol))
nrow(diamond_gene)
```
```{r}
sum(!is.na(res_annot$stitle))
```

#Significant flutamide_effect_F103vsC103
```{r}
res_sig <- subset(
  res,
  padj < 0.05 & abs(log2FoldChange) > 1 
)

sig_annot <- subset(
  res_annot,
  padj < 0.05 &  !is.na(res_annot$gene_symbol) & abs(log2FoldChange) > 1 
)

nrow(res_sig)
nrow(sig_annot)

write.csv(
  sig_annot,
  "sig_annot_F103vsC103.csv",
  row.names = FALSE
)
```

```{r}
# Volcano plot with gene labels
res_df <- as.data.frame(res_annot) %>%
  mutate(
    significant = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1,
    direction   = case_when(
      significant & log2FoldChange > 1  ~ "up",
      significant & log2FoldChange < -1 ~ "down",
      TRUE ~ "ns"
    ),
    # only label significant genes that have a gene symbol
    label = ifelse(significant & !is.na(gene_symbol), gene_symbol, NA)
  )

volcano <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj),
                               color = direction)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_text_repel(
    aes(label = label),
    size = 2.5,
    max.overlaps = 20,
    na.rm = TRUE,
    box.padding = 0.3,
    segment.color = "grey50",
    show.legend = FALSE 
  ) +
  scale_color_manual(values = c("up"   = "#D6604D",
                                "down" = "#2166AC",
                                "ns"   = "grey70")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1),     linetype = "dashed") +
  theme_bw() +
  theme(legend.position = "none") +    # <-- add this
  xlab("log2 Fold Change") +
  ylab("-log10 adjusted p-value")

ggsave(file.path(base_dir, "figures", "F103_C103_volcano_labeled.png"),
       volcano, width = 4, height = 5)
```


```{r}
head(
  sig_annot[order(sig_annot$padj),
            c("gene_id","log2FoldChange","padj","gene_symbol","stitle")],
  30
)

gene_symbols_up_down_F103vsC103 <- unique(sig_annot$gene_symbol)

write.table(
  data.frame(gene_symbol = gene_symbols_up_down_F103vsC103),
  file = "up_down_F103vsC103_symbols.tsv",
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE,
  quote = FALSE
)

```
```{r}
# search for androgen receptor in all genes regardless of significance
ar_genes <- res_annot[
  grepl("androgen", res_annot$stitle, ignore.case = TRUE) |
  grepl("\\bAR\\b", res_annot$gene_symbol, ignore.case = FALSE) |
  grepl("androgen", res_annot$gene_symbol, ignore.case = TRUE),
]

nrow(ar_genes)

ar_genes[, c("gene_id", "log2FoldChange", "padj", "gene_symbol", "stitle")]
```

```{r}
# check known AR target genes
ar_targets <- res_annot[
  grepl("kallikrein|KLK|PSA|NKX3|FKBP5|TMPRSS2|SGK1|PPAP2a|
         UGT2B|SARG|NDRG1|KLF", 
        res_annot$gene_symbol, ignore.case = TRUE),
]

ar_targets[, c("gene_id", "log2FoldChange", "padj", "gene_symbol", "stitle")]
```


# body 
```{r}
# ============================================================
# BODY SAMPLES ONLY — F103 vs C103
# ============================================================

dds_body <- dds[, dds$tissue == "body"]
dds_body$group <- droplevels(dds_body$group)
design(dds_body) <- ~ group

keep_body <- rowSums(counts(dds_body)) >= 10
dds_body <- dds_body[keep_body, ]
dds_body <- DESeq(dds_body)

res_flut_F103vsC103_BODY <- run_contrast(
  dds_body,
  contrast = c("group", "flutamide_day103", "control_day103"),
  name     = "flutamide_F103vsC103_BODY",
  base_dir = base_dir
)
```

```{r}
# Volcano plot with gene labels
res_df <- as.data.frame(res_res_flut_F103vsC103_BODY_annot) %>%
  mutate(
    significant = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1,
    direction   = case_when(
      significant & log2FoldChange > 1  ~ "up",
      significant & log2FoldChange < -1 ~ "down",
      TRUE ~ "ns"
    ),
    # only label significant genes that have a gene symbol
    label = ifelse(significant & !is.na(gene_symbol), gene_symbol, NA)
  )

volcano <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj),
                               color = direction)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_text_repel(
    aes(label = label),
    size = 2.5,
    max.overlaps = 20,
    na.rm = TRUE,
    box.padding = 0.3,
    segment.color = "grey50",
    show.legend = FALSE 
  ) +
  scale_color_manual(values = c("up"   = "#D6604D",
                                "down" = "#2166AC",
                                "ns"   = "grey70")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1),     linetype = "dashed") +
  theme_bw() +
  theme(legend.position = "none") +    # <-- add this
  xlab("log2 Fold Change") +
  ylab("-log10 adjusted p-value")

ggsave(file.path(base_dir, "figures", "F103_C103_BODY_volcano_labeled.png"),
       volcano, width = 2, height = 2)

```


#head 
```{r}
# ============================================================
# HEAD SAMPLES ONLY — F103 vs C103
# ============================================================

dds_head <- dds[, dds$tissue == "head"]
dds_head$group <- droplevels(dds_head$group)
design(dds_head) <- ~ group

keep_head <- rowSums(counts(dds_head)) >= 10
dds_head <- dds_head[keep_head, ]
dds_head <- DESeq(dds_head)

res_flut_F103vsC103_HEAD <- run_contrast(
  dds_head,
  contrast = c("group", "flutamide_day103", "control_day103"),
  name     = "flutamide_F103vsC103_HEAD",
  base_dir = base_dir
)


```

```{r}
res_flut_F103vsC103_BODY <- read.csv(
  "results/flutamide_F103vsC103_BODY_DESeq2_results.csv",
  row.names = 1,
  stringsAsFactors = FALSE
)

res_flut_F103vsC103_BODY$gene_id <- rownames(res_flut_F103vsC103_BODY)

res_res_flut_F103vsC103_BODY_annot <- merge(res_flut_F103vsC103_BODY, annot, by = "gene_id", all.x = TRUE)

sig_res_res_flut_F103vsC103_BODY_annot <- subset(
  res_res_flut_F103vsC103_BODY_annot,
  padj < 0.05 &  !is.na(res_res_flut_F103vsC103_BODY_annot$gene_symbol) & abs(log2FoldChange) > 1 
)

nrow(sig_res_res_flut_F103vsC103_BODY_annot)

head(
  sig_res_res_flut_F103vsC103_BODY_annot[order(sig_res_res_flut_F103vsC103_BODY_annot $padj),
            c("gene_id","log2FoldChange","padj","gene_symbol","stitle")],
  30
)

gene_symbols_res_flut_F103vsC103_BODY_annot <- unique(sig_res_res_flut_F103vsC103_BODY_annot$gene_symbol)

write.table(
  data.frame(gene_symbol = gene_symbols_res_flut_F103vsC103_BODY_annot),
  file = "res_flut_F103vsC103_BODY.tsv",
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE,
  quote = FALSE
)
```

## HEAD
```{r}
res_flut_F103vsC103_HEAD <- read.csv(
  "results/flutamide_F103vsC103_HEAD_DESeq2_results.csv",
  row.names = 1,
  stringsAsFactors = FALSE
)

res_flut_F103vsC103_HEAD$gene_id <- rownames(res_flut_F103vsC103_HEAD)

res_res_flut_F103vsC103_HEAD_annot <- merge(res_flut_F103vsC103_HEAD, annot, by = "gene_id", all.x = TRUE)

sig_res_res_flut_F103vsC103_HEAD_annot <- subset(
  res_res_flut_F103vsC103_HEAD_annot,
  padj < 0.05 &  !is.na(res_res_flut_F103vsC103_HEAD_annot$gene_symbol) & abs(log2FoldChange) > 1 
)

nrow(sig_res_res_flut_F103vsC103_HEAD_annot)

head(
  sig_res_res_flut_F103vsC103_HEAD_annot[order(sig_res_res_flut_F103vsC103_HEAD_annot $padj),
            c("gene_id","log2FoldChange","padj","gene_symbol","stitle")],
  30
)

gene_symbols_res_flut_F103vsC103_HEAD_annot <- unique(sig_res_res_flut_F103vsC103_HEAD_annot$gene_symbol)

write.table(
  data.frame(gene_symbol = gene_symbols_res_flut_F103vsC103_HEAD_annot),
  file = "res_flut_F103vsC103_HEAD.tsv",
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE,
  quote = FALSE
)

write.csv(
  sig_res_res_flut_F103vsC103_HEAD_annot,
  "sig_annot_F103vsC103_HEAD.csv",
  row.names = FALSE
)

write.csv(
  sig_res_res_flut_F103vsC103_BODY_annot,
  "sig_annot_F103vsC103_BODY.csv",
  row.names = FALSE
)

###
res_flut_F103vsC103_HEAD_sig <- subset(
  res_flut_F103vsC103_HEAD,
  padj < 0.05 & abs(log2FoldChange) > 1 
)

nrow(res_flut_F103vsC103_HEAD_sig)
nrow(sig_res_res_flut_F103vsC103_HEAD_annot)

###
res_flut_F103vsC103_BODY_sig <- subset(
  res_flut_F103vsC103_BODY,
  padj < 0.05 & abs(log2FoldChange) > 1 
)

nrow(res_flut_F103vsC103_BODY_sig)
nrow(sig_res_res_flut_F103vsC103_BODY_annot)



```

```{r}
# Volcano plot with gene labels
res_df <- as.data.frame(res_res_flut_F103vsC103_HEAD_annot) %>%
  mutate(
    significant = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1,
    direction   = case_when(
      significant & log2FoldChange > 1  ~ "up",
      significant & log2FoldChange < -1 ~ "down",
      TRUE ~ "ns"
    ),
    # only label significant genes that have a gene symbol
    label = ifelse(significant & !is.na(gene_symbol), gene_symbol, NA)
  )

volcano <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj),
                               color = direction)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_text_repel(
    aes(label = label),
    size = 2.5,
    max.overlaps = 20,
    na.rm = TRUE,
    box.padding = 0.3,
    segment.color = "grey50",
    show.legend = FALSE 
  ) +
  scale_color_manual(values = c("up"   = "#D6604D",
                                "down" = "#2166AC",
                                "ns"   = "grey70")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1),     linetype = "dashed") +
  theme_bw() +
  theme(legend.position = "none") +    # <-- add this
  xlab("log2 Fold Change") +
  ylab("-log10 adjusted p-value")

ggsave(file.path(base_dir, "figures", "F103_C103_HEAD_volcano_labeled.png"),
       volcano, width = 2, height = 2)
```


```{r}
head(
  sig_res_res_flut_F103vsC103_HEAD_annot[
    order(sig_res_res_flut_F103vsC103_HEAD_annot$padj),
    c("gene_id","log2FoldChange","padj","gene_symbol","stitle")],
  33
)
```


```{r}
# go back to the pre-subset annotated dataframe
all_sig_head <- subset(
  res_res_flut_F103vsC103_HEAD_annot,  # this is BEFORE the gene_symbol filter
  padj < 0.05 & abs(log2FoldChange) > 1
)

nrow(all_sig_head)  # should be 33

head(
  all_sig_head[order(all_sig_head$padj),
               c("gene_id","log2FoldChange","padj","gene_symbol","stitle")],
  33
)
```

# Final HEAD BODY tablestables

```{r}

res_flut_F103vsC103_HEAD <- read.csv(
  "results/flutamide_F103vsC103_HEAD_DESeq2_results.csv",
  row.names = 1,
  stringsAsFactors = FALSE
)

res <- read.csv(
  "results/flutamide_effect_F103vsC103_DESeq2_results.csv",
  row.names = 1,
  stringsAsFactors = FALSE
)

res$gene_id <- rownames(res)

annot <- read.csv(
  "diamond_gene_annotation_v1.csv",
  stringsAsFactors = FALSE
)

res_annot <- merge(res, annot, by = "gene_id", all.x = TRUE)

write.csv(
  res_annot,
  "flutamide_effect_F103vsC103_DESeq2_results_annotated.csv",
  row.names = FALSE
)
```

```{r}
sum(!is.na(diamond_gene$gene_symbol))
nrow(diamond_gene)
```
```{r}
sum(!is.na(res_annot$stitle))
```

#Significant flutamide_effect_F103vsC103
```{r}
res_sig <- subset(
  res,
  padj < 0.05 & abs(log2FoldChange) > 1 
)

sig_annot <- subset(
  res_annot,
  padj < 0.05 &  !is.na(res_annot$gene_symbol) & abs(log2FoldChange) > 1 
)

nrow(res_sig)
nrow(sig_annot)

write.csv(
  sig_annot,
  "sig_annot_F103vsC103.csv",
  row.names = FALSE
)
